Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-05 doc/invt.tex
figures: improve sinc.gp

We now adapt the above formulae to a non-periodic variable.
%
We shall use the reflective boundary condition\cite{bussi2006},
and modifies the form of the updating matrix, $\mathbf w$, as
%
%
\begin{equation}
  w_{ij}
  =
  \mu_{ i - j }
  +
  \mu_{ i + j - 1 }
  +
  \mu_{ i + j - 2 n - 1 },
  \label{eq:w_band}
\end{equation}
%
Again the updating kernel satisfies
Eqs. \eqref{eq:mu_normalization}
and
\eqref{eq:mu_symm}.
%
However,
the radius $b$ only need to be less than $n$
instead of $n/2$.
%
The last two terms
on the right-hand side of \eqref{eq:w_band} are added
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006},
such that Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance}
are satisfied.

To compute the eigenvectors,
we shall replace Eq. \eqref{eq:phi_pbc}
and redefine the out-of-boundary values
of $\varphi_j$ as
%
\begin{equation}
  \varphi_j
  =
  \begin{dcases}
    \varphi_{ 1 - j }           & \mathrm{for \;} j \le 0, \\
    \varphi_{ 2 \, n + 1 - j }  & \mathrm{for \;} j > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
One can show that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \varphi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \varphi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \varphi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \varphi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \varphi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \varphi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \varphi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \varphi_j.
  \end{aligned}
  $$
}
%
Thus, it is also possible to construct
$n$ orthonormal eigenvectors,
$\pmb\varphi^{(1)}, \dots, \pmb\varphi^{(n)}$,
as
%
\begin{equation}
\varphi^{(k)}_i
=
\varphi_{k i}
=
\sqrt{
    \frac{ 2 - \delta_{k, 1} }
         {       n           }
     }
   \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                  n                         }.
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  \label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{We derive the above two equations below.

@@

\subsubsection{\label{sec:optscheme}
Brick-wall band-pass updating schemes}

We can also generalize
the single-bin scheme to a class of
\emph{brick-wall band-pass} updating schemes
that allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the error modes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is changed to
%
\begin{equation}
  E_A
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

@@

In the above, we have been focusing on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part,
Eq. \eqref{eq:error_res}
and minimize the total error.
%
For the brick-wall band-pass updating schemes,
this allows to determine the optimal value
of the free parameter, $t_0$.

@@

Again, if we are certain that
the intrinsic PMF is sufficiently smooth,
then we can explicitly filter some error modes,
say those for $k > K$,
in the histogram, and the sum in
Eq. \eqref{eq:error_equil}
can be truncated at $k = K$.

@@

Our test system is
a model one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$
under the reflective boundary condition.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
For convenience,
the intrinsic distribution, $p^*_i$,
is a set to be flat as well.
%
Our goal is to observe how the fluctuating error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule, $\alpha(t)$,
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and equilibrate the system under
a constant updating magnitude,
$\alpha(t) = \alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the target schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

We first tested the theory using
the inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a parameter to be optimized.
and $t_0 = 1/(\lambda \, \alpha_0)$
to smooth the transition from equilibration.
%
This schedule is analytically simpler
than the optimal schedule,
and it is studied in Appendix \ref{sec:invt_schedule}.
%
We wish to see how the final error
depends on the parameter, $\lambda$,
and if it matches the predicted value,
given by Eq. \eqref{eq:error_invt}.

@@

We also observe
that the two error functions
differed only by a multiplicative constant,
again again agreeing with the theory.
%
With identical eigenvalues,
$\lambda_1 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling process only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

@@

In Fig. \ref{fig:err_sigscan},
we compared the error from the inverse-time schedule
with the optimal $\lambda$,
and that from the asymptotically optimal schedule,
for the Gaussian updating scheme.
%
The errors from the two schedules appeared to be reasonable close,
unless for large width case.
%
This suggests that very long metadynamics simulations
may benefit from optimal schedule,
instead of the simple formula, given by
Eq. \eqref{eq:alpha_invtlambda}.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:err_sigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width of the Gaussian updating scheme,
    $\sigma$,
    according to the schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal $\lambda$,
    and or the asymptotically optimal schedule.
  }
\end{center}
\end{figure}

%\subsection{Optimality of the single-bin scheme}

The above comparison also shows the asymptotic superiority
of the single-bin scheme.
It is clear that for short times,
the single-bin (WL) may not be optimal,
but it ultimately wins out as simulation lengthens.

%Randomize the initial error?
%
%Error vs. $\mu_1$.
%
%Error vs. $\sigma$.

@@

We have also compared different updating schemes.
%
We have shown that in the long time limit
the single-bin scheme belongs to a class of
asymptotically optimal updating schemes.
%
%That is, the single-bin scheme
%is asymptotically as efficient as
%an equilibrium FDS simulation
%with the exact bias potential.

@@

Real simulations, however,
can be relatively short
and need not enter
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(T) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ T + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      -
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

@@

Note that Eq. \eqref{eq:lambda_nn_onestep}
is an approximation good only
for sufficiently short simulations.
%
For very longer simulations,
the optimal $\lambda$
will ultimately approach
the smallest eigenvalue,
$\min \lambda_k = 1 - 4 \, \mu_1$
according to the discussion in Sec. \ref{sec:optlambda}.

>>>>

To adapt the above formulae to a non-periodic variable,
we use the reflective boundary condition\cite{bussi2006},
and change the form of the updating matrix to
%
%
\begin{equation}
  w_{ij}
  =
  \mu_{ i - j }
  +
  \mu_{ i + j - 1 }
  +
  \mu_{ i + j - 2 n - 1 },
  \label{eq:w_band_refl}
\end{equation}
%
Again the updating kernel satisfies
Eqs. \eqref{eq:mu_normalization}
and \eqref{eq:mu_symm}.
%
However,
the radius $b$ only need to be less than $n$.
%
The last two terms
on the right-hand side of \eqref{eq:w_band_refl} are added
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006},
such that Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance}
are satisfied.

Similar to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\varphi_j$ as
%
\begin{equation}
  \varphi_j
  =
  \begin{dcases}
    \varphi_{ 1 - j }           & \mathrm{for \;} j \le 0, \\
    \varphi_{ 2 \, n + 1 - j }  & \mathrm{for \;} j > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \varphi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \varphi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \varphi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \varphi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \varphi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \varphi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \varphi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \varphi_j.
  \end{aligned}
  $$
}
%
Thus, we also have $n$ orthonormal eigenvectors,
$\pmb\varphi^{(1)}, \dots, \pmb\varphi^{(n)}$,
as
%
\begin{equation}
\varphi^{(k)}_i
=
\varphi_{k i}
=
\sqrt{
    \frac{ 2 - \delta_{k, 1} }
         {       n           }
     }
   \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                  n                         }.
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  \label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{We derive the above two equations below.

@@

\subsubsection{\label{sec:optscheme}
Bandpass updating schemes}

We can also generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the error modes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is changed to
%
\begin{equation}
  E_A
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

@@

In the above, we have been focusing on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part,
Eq. \eqref{eq:error_res}
and minimize the total error.
%
For the bandpass updating schemes,
this allows to determine the optimal value
of the free parameter, $t_0$.

@@

Again, the above argument can be extended
to the bandpass updating schemes.
%
If we are certain that
the intrinsic PMF is sufficiently smooth,
then we can explicitly filter some error modes,
say those for $k > K$,
in the histogram, and the sum in
Eq. \eqref{eq:error_equil}
can be truncated at $k = K$.

@@

Our test system is
a model one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$.
%
Unless specified otherwise,
we assume $i$ is a non-periodic variable
under the reflective boundary condition.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
For convenience,
the intrinsic distribution, $p^*_i$,
is a set to be flat as well.
%
Our goal is to observe how the fluctuating error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule, $\alpha(t)$,
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and equilibrate the system under
a constant updating magnitude,
$\alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the target schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

We first tested the theory using
the inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a parameter to be optimized
with $t_0 = 2/(\lambda \, \alpha_0)$.
%
This simple schedule
allows us to predict the error analytically
[cf. Appendix \ref{sec:invt_schedule}],
and it serves as a model of verifying our theory.

@@

The two error functions
differed only by a multiplicative constant,
again agreeing with the theory.
%
With identical eigenvalues,
$\lambda_1 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling process only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

While the optimal schedules of the Gaussian updating schemes
are rather complex,
those of the bandpass updating schemes
are always given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
with the optimal $t_0$
given by Eq. \eqref{eq:t0_sinc},
as shown in Fig. \ref{fig:sinc}(a).

In Fig. \ref{fig:sinc}(b),
we show that the updating kernels
of the bandpass updating schemes
are similar to Gaussian ones
except the oscillatory tails.
%
The equivalent width, $\sigma$, is roughly given
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  ,
\label{eq:sigma_equiv}
\end{equation}
where $g = 2$ for the periodic case
or $g = 1$ otherwise.
%
The kernel is a zigzag in the non-periodic case
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, the ruggedness disappears
in the actual updating matrix, Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by either the term for $l = i + j - 1$
or that for $l = i + j - 2 \, n - 1$,
as shown in the inset
of Fig. \ref{fig:sinc}(b).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    (a) Optimal schedules of the bandpass
    updating schemes for a periodic variable
    with $K = 4$,
    and for a non-periodic variable
    with $K = 8$.
    (b) Corresponding updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix according to Eq. \eqref{eq:w_band_refl}.
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
the optimal updating scheme
depends on the simulation length, $T$,
and other parameters.
%
To illustrate this,
we compare Gaussian updating schemes
with different width, $\sigma$.
%
For the global sampling process,
we see from Fig. \ref{fig:errsigscan}(a)
that the lowest error was achieved
by the widest window for the shorter simulation
of $T = 10^8$,
while the minimum moved to $\sigma = 0$
as the simulation lengthens to $T = 10^{10}$.
%
However,
the shift was hard to detect
for the local sampling scheme,
as shown in Fig. \ref{fig:errsigscan}(b).
%
In all cases,
the bandpass scheme
appeared to be a superior
alternative to the Gaussian updating scheme.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the brick-wall bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
  }
\end{center}
\end{figure}

@@

We have also compared different updating schemes.
%
We have shown that in the long time limit
the single-bin scheme belongs to a class of
brick-wall bandpass updating schemes that are
asymptotically optimal.
%
%That is, the single-bin scheme
%is asymptotically as efficient as
%an equilibrium FDS simulation
%with the exact bias potential.

@@

Real simulations, however,
can be relatively short
and need not enter
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(T) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ T + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that
the system was initially equilibrated
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql},
as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      -
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

@@

Note that Eq. \eqref{eq:lambda_nn_onestep}
is an approximation good only
for sufficiently short simulations,
since we have taken the limit of $n \to \infty$
for a fixed $T$.
%
For very long simulations,
the optimal $\lambda$
will ultimately approach
the smallest eigenvalue,
$\min \lambda_k = 1 - 4 \, \mu_1$
according to the discussion in Sec. \ref{sec:optlambda}.
